---
title: "Marine biodiversity data ingestion for OBIS (DwC translation)"
subtitle: "Long term monitoring of fish abundances from coastal Skagerrak"
author:
- [your name here]
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/iobis/mpaeu_di_fish_abund_skagerrak/blob/master/src/obisdi_general.Rmd).

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)      # Data manipulation
library(obisdi)         # Tools for data ingestion for OBIS
library(here)           # Get paths (important!)
library(arrow)          # To deal with parquet files
library(worrms)         # Taxonomy checking
```

# Read source data

This dataset was downloaded from the BioTIME^[https://biotime.st-andrews.ac.uk/] database and can be found in this link: https://biotime.st-andrews.ac.uk/selectStudy.php?study=428. Two files are available on this link: the raw data and the metadata.

```{r}
# Get the path to data/raw
raw_path <- here("data", "raw")

list.files(raw_path)
```

# Preprocessing

We first read the raw data:

```{r}
dataset <- read.csv(paste0(raw_path, "/raw_data_428.csv"))

str(dataset)

head(dataset)
```

By reading the `metadata` we can see that there is no biomass information, so we are interested in the abundance information:

```{r include=FALSE}
metadata <- read.csv(paste0(raw_path, "/metadata_428.csv"))
```

> `r metadata$DATA_SOURCE[1]`

We select only the relevant collumns. We also add a unique ID to each occurrence to be able to track back any transformation on data.

```{r}
dataset <- select(dataset, -BIOMAS, -SAMPLE_DESC)

dataset$uniqueID <- paste0("biotime_study248_", 1:nrow(dataset))
```

# Darwin Core mapping

We will need to generate two files for the submission:
1. event - describe the sampling event
2. occurrence - the occurrence of species in an event

## `event` table

We start by getting each unique event. Both the occurrences and the events table need to have the same eventID and locationID, so we still work with the `dataset` object.

```{r}
dataset <- dataset %>%
  arrange(-DAY, -MONTH, -YEAR)

dataset <- dataset %>%
  group_by(LATITUDE, LONGITUDE, PLOT, DAY, MONTH, YEAR) %>%
  mutate(dwc_eventID = paste0("skag_ev_", cur_group_id())) %>%
  group_by(LATITUDE, LONGITUDE, PLOT) %>%
  mutate(dwc_locationID = paste0("skag_lo_", cur_group_id()))
```

Correct the name of Longitude/Latitude fields, and date.

```{r}
dataset <- dataset %>%
  mutate(dwc_decimalLatitude = LATITUDE,
         dwc_decimalLongitude = LONGITUDE) %>%
  mutate(dwc_eventDate = paste(YEAR, MONTH, DAY, sep = "-")) %>%
  mutate(dwc_day = DAY, dwc_year = YEAR, dwc_month = MONTH)
```

Now we create a new object called `events` that will be our events table. We add other columns that are needed:

```{r}
events <- dataset %>%
  mutate(dwc_type = "Event",
         dwc_ownerInstitutionCode = "HAVFORSKNINGSINSTITUTTET",
         dwc_samplingProtocol = "Beach seine haul (40m long, mesh size 1.5cm), nearshore (<15m depth) - all species counted.",
         dwc_sampleSizeUnit = "square metre",
         dwc_sampleSizeValue = "5-700",
         dwc_samplingEffort = "transect",
         dwc_country = "Norway",
         dwc_countryCode = "NO",
         dwc_locality = "Skagerrak",
         dwc_datasetID = paste0("biotime_study", metadata$STUDY_ID[1]),
         dwc_datasetName = "Long term monitoring of fish abundances from coastal Skagerrak.")
```

Get only the unique events:

```{r}
events <- events %>%
  ungroup() %>%
  distinct(dwc_eventID, .keep_all = T)
```

## `occurrence` table

We first create a new object that will have our occurrences:

```{r}
occurrences <- dataset
```

Then we add the needed columns:

```{r}
occurrences <- occurrences %>%
  mutate(dwc_type = "Event",
         dwc_ownerInstitutionCode = "HAVFORSKNINGSINSTITUTTET",
         dwc_occurrenceID = uniqueID,
         dwc_basisOfRecord = "HumanObservation",
         #dwc_scientificName = GENUS_SPECIES, # We will add it later, corrected
         dwc_individualCount = ABUNDANCE,
         dwc_organismQuantity = ABUNDANCE,
         dwc_organismQuantityType = "individuals")
```

We have no information if this dataset was checked with WoRMS for consistency of taxonomy. Thus, we do a check here:

```{r}
# Look into WoRMS for each name
name_checking <- wm_records_names(unique(occurrences$GENUS_SPECIES))

# Verify if there is any record with not valid names
table(unlist(lapply(name_checking, function(x){x$status})))

# We see that there are three unaccepted records. For these, we will extract the correct names.
# We also check if there are records for which we were unable to find information:
no_match <- unique(occurrences$GENUS_SPECIES)[unlist(lapply(name_checking, nrow)) == 0]
no_match
```

We have some records that needs updated names and also 4 records for which there was no match. One of them is a unknown taxa and one is a non marine species^[The species _Polioptila caerulea_ returns a bird in all our searches. It's possible that this is a mistake when the name was written in the table.] both which we will exclude, while for the other 2 we manually correct the information:

```{r}
name_checking <- lapply(name_checking, function(x){
  if (nrow(x) == 0) {
    x <- name_checking[[1]][1,] 
    x[1,] <- NA
  }
  x
})

name_info <- bind_rows(name_checking)

name_info <- name_info %>%
  select(valid_name, valid_AphiaID, valid_authority,
                kingdom, phylum, class, order, family, genus, rank) %>%
  mutate(scientificName = valid_name,
         scientificNameAuthorship = valid_authority,
         scientificNameID = paste0("urn:lsid:marinespecies.org:taxname:", valid_AphiaID),
         taxonRank = rank) %>%
  select(-valid_name, -valid_authority, -valid_AphiaID, -rank)

name_info$GENUS_SPECIES <- unique(occurrences$GENUS_SPECIES)

occurrences <- occurrences %>%
  filter(!GENUS_SPECIES %in% c("Manetyngel unknown", "Polioptila caerulea"))
name_info <- name_info %>%
  filter(!GENUS_SPECIES %in% c("Manetyngel unknown", "Polioptila caerulea"))

name_info[name_info$GENUS_SPECIES == "Sygnathus typhle",1:10] <-
  list("Animalia", "Chordata", "Teleostei", "Syngnathiformes", "Syngnathidae", "Syngnathus", "Syngnathus typhle", "(Linnaeus, 1758)",
       "urn:lsid:marinespecies.org:taxname:127393", "Species")

name_info[name_info$GENUS_SPECIES == "Labrus bimaculatus",1:10] <-
  list("Animalia", "Chordata", "Teleostei", NA, "Labridae", "Labrus", "Labrus mixtus", "(Linnaeus, 1758)",
       "urn:lsid:marinespecies.org:taxname:151501", "Species") # According to FishBase

# Merge correct names
colnames(name_info)[1:10] <- paste0("dwc_", colnames(name_info)[1:10])

# Add originalNameUsage for the one we searched on FishBase
name_info$dwc_originalNameUsage <- NA
name_info$dwc_originalNameUsage[name_info$GENUS_SPECIES == "Labrus bimaculatus"] <-
  "Labrus bimaculatus"

occurrences <- left_join(occurrences, name_info, by = "GENUS_SPECIES")
occurrences$dwc_taxonRank <- tolower(occurrences$dwc_taxonRank)
```

Now we check if both tables have the same number of unique `locationID` and `eventID`:

```{r}
all.equal(unique(occurrences$dwc_eventID), unique(occurrences$dwc_eventID))

all.equal(unique(occurrences$dwc_locationID), unique(occurrences$dwc_locationID))
```

# Post-processing

Remove unused columns and change names of those with *dwc_*:

```{r}
events <- events %>%
  mutate(dwc_id = uniqueID,
         dwc_eventRemarks = paste("Plot", PLOT)) %>%
  select(starts_with("dwc_"))

occurrences <- occurrences %>%
  mutate(dwc_id = uniqueID) %>%
  ungroup() %>%
  select(starts_with("dwc_"))

# Change column names
colnames(events) <- str_remove(colnames(events), "dwc_")

colnames(occurrences) <- str_remove(colnames(occurrences), "dwc_")

# Remove columns from the occurrences table
occurrences <- occurrences %>%
  select(-day, -month, -year, -decimalLatitude, -decimalLongitude, -eventDate,
         -locationID)
```

We now have our final files:

```{r}
occurrences

events
```


# Export final files

We export in `csv` format:

```{r}
proc_path <- here("data", "processed")

write_csv(events, paste0(proc_path, "/events.csv"))
write_csv(occurrences, paste0(proc_path, "/occurrences.csv"))
```

